Ten Simple Rules for Teaching Programming

### 2018-02-23 paper.tex
Revising
There are always more changes

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Ten quick tips for teaching programming} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
{Neil C.C.\ Brown}\textsuperscript{1{\ddag}*},
{Greg~Wilson}\textsuperscript{2{\ddag}*}
\\
\bigskip
\textbf{1} King's College London / neil.c.c.brown@kcl.ac.uk \\
\textbf{2} DataCamp / gvwilson@third-bit.com \\

@@

\item The instructor then gives learners a multiple choice question
  that probes for misconceptions rather than simple factual recall.
  A programming example is given in Figure~\ref{fig:mcq}
  which relates to integer comparison and loops.  The multiple choice question must be well designed.
  There is no point asking a trivial question that all students will get right or one with meaningless wrong answers which no student will pick.
  The ideal questions are those where 40--60\% of students are likely to get the right answer first time~\cite[p23]{nrc-reaching}, and where every wrong answer corresponds to a misconception that will cause it to be picked by at least some students.

@@

Note that live coding does not always have to start with a blank screen:
instructors may give students some starter code
that relies solely on concepts they have already mastered,
and then extend it or modify it with live coding.

@@

Margulieux and Morrison et al.~\cite{morrison-subgoals,morrison-parsons,margulieux} have shown that students perform better
when worked examples are broken down into steps (or subgoals)
which are given names (or labels) -- an example is given in Figure~\ref{fig:subgoal}.
Subgoal labels provide a structure which allow learners to see the similarities between coding problems and to communicate with their peers and instructors more efficiently.  Learners can then apply the labels to future tasks that they attempt themselves.

@@

We are grateful to Barbara Ericson and Leo Porter for their comments on a draft of this article, and to Lauren Margulieux for providing the example used in Figure \ref{fig:subgoal}.

@@

\begin{figure}[h]
\begin{mdframed}
\begin{verbatim}
for (int i = 1; i < 10; i++) {
    if (i < 3 || i >= 8) {
        System.out.println("Yes");
    }
}
\end{verbatim}
How many times will the above code print out the word Yes?\\
a) 10\\
b) 5\\
c) 4\\
d) 3
\end{mdframed}
\caption{An example multiple choice question probing learners' understanding of loops and integer comparisons.}
\label{fig:mcq}
\end{figure}

\begin{figure}[h]
\begin{mdframed}
\noindent
\textbf{Conventional Materials}
\begin{verbatim}
1. Click on "AccelerometerSensor1"
2. Drag out a when AccelerometerSensor1.AccelerationChanged block
3. Click on ?cowbellSound? 
4. Drag out call cowbellSound.Play and connect it after
      when AccelerometerSensor1.AccelerationChanged
\end{verbatim}

@@

Set Output from My Blocks
3. Click on ?cowbellSound? 
4. Drag out call cowbellSound.Play and connect it after
      when AccelerometerSensor1.AccelerationChanged
\end{verbatim}
\end{mdframed}
\caption{An example of subgoal labeling.}
\label{fig:subgoal}
\end{figure}

@@

\bibliography{paper}

>>>>

% Trying to typeset code
\usepackage{newfloat}
\DeclareFloatingEnvironment[
    fileext=loc,
    listname={List of Code Samples},
    name=Code,
    placement=tbhp,
    within=none,
]{code}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Ten quick tips for teaching programming} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
{Neil C.C.\ Brown}\textsuperscript{1{\ddag}},
{Greg~Wilson}\textsuperscript{2{\ddag}*}
\\
\bigskip
\textbf{1} Department of Informatics, King's College London, London, UK, neil.c.c.brown@kcl.ac.uk \\
\textbf{2} DataCamp, Toronto, Onario, Canada, gvwilson@third-bit.com \\

@@

\item The instructor then gives learners a multiple choice question
  that probes for misconceptions rather than simple factual recall.
  A programming example is given in Code~\ref{fig:mcq}
  which relates to integer comparison and loops.  The multiple choice question must be well designed.
  There is no point asking a trivial question that all students will get right or one with meaningless wrong answers which no student will pick.
  The ideal questions are those where 40--60\% of students are likely to get the right answer first time~\cite[p23]{nrc-reaching}, and where every wrong answer corresponds to a misconception that will cause it to be picked by at least some students.

@@

Note that live coding does not always have to start with a blank screen:
instructors may give students some starter code
that relies solely on concepts they have already mastered,
and then extend it or modify it with live coding.
Instructors who use live coding should ensure that
learners have reference material available after lectures,
such as a textbook,
but should also recognize that students of all ages
increasingly turn to Q\&A sites such as Stack Overflow for information.

@@

Margulieux and Morrison et al.~\cite{morrison-subgoals,morrison-parsons,margulieux} have shown that students perform better
when worked examples are broken down into steps (or subgoals)
which are given names (or labels) -- an example is given in Code~\ref{fig:subgoal}.
Subgoal labels provide a structure which allow learners to see the similarities between coding problems and to communicate with their peers and instructors more efficiently.  Learners can then apply the labels to future tasks that they attempt themselves.

@@

We are grateful to Barbara Ericson and Leo Porter for their comments on a draft of this article,
and to Lauren Margulieux for providing the example used in Code~\ref{fig:subgoal}.

@@

\begin{code}[h]
\begin{mdframed}
\begin{verbatim}
for (int i = 1; i < 10; i++) {
    if (i < 3 || i >= 8) {
        System.out.println("Yes");
    }
}
\end{verbatim}
How many times will the above code print out the word Yes?\\
a) 10\\
b) 5\\
c) 4\\
d) 3
\end{mdframed}
\caption{An example multiple choice question probing learners' understanding of loops and integer comparisons.}
\label{fig:mcq}
\end{code}

\begin{code}[h]
\begin{mdframed}
\noindent
\textbf{Conventional Materials}
\begin{verbatim}
1. Click on "AccelerometerSensor1"
2. Drag out a when AccelerometerSensor1.AccelerationChanged block
3. Click on "cowbellSound"
4. Drag out call cowbellSound.Play and connect it after
   when AccelerometerSensor1.AccelerationChanged
\end{verbatim}

@@

Set Output from My Blocks
3. Click on "cowbellSound"
4. Drag out call cowbellSound.Play and connect it after
   when AccelerometerSensor1.AccelerationChanged
\end{verbatim}
\end{mdframed}
\caption{An example of subgoal labeling.}
\label{fig:subgoal}
\end{code}

@@

% \bibliography{paper}

\begin{thebibliography}{10}

\bibitem{mayer-subject}
Mayer RE.
\newblock Teaching of Subject Matter.
\newblock Annual Review of Psychology. 2004;55(1):715--744.
\newblock doi:{10.1146/annurev.psych.55.082602.133124}.

\bibitem{cs-teaching-tips}
Lewis C. CS Teaching Tips; 2017.
\newblock \texttt{http://csteachingtips.org/}.

\bibitem{guzdial-myths}
Guzdial M. Top 10 Myths About Teaching Computer Science; 2015.
\newblock
  \texttt{https://cacm.acm.org/blogs/blog-cacm/189498-top-10-myths-about-teaching-computer-science/fulltext}.

\bibitem{patitsas-bimodal}
Patitsas E, Berlin J, Craig M, Easterbrook S.
\newblock Evidence That Computer Science Grades Are Not Bimodal.
\newblock In: Proceedings of the 2016 ACM Conference on International Computing
  Education Research. ICER '16. New York, NY, USA: ACM; 2016. p. 113--121.
\newblock Available from: \url{http://doi.acm.org/10.1145/2960310.2960312}.

\bibitem{alvidrez-weinstein}
Alvidrez J, Weinstein RS.
\newblock Early teacher perceptions and later student academic achievement.
\newblock Journal of Educational Psychology. 1999;91(4):731--746.

\bibitem{brophy}
Brophy JE.
\newblock Research on the self-fulfilling prophecy and teacher expectations.
\newblock Journal of Educational Psychology. 1983;75(5):631--661.

\bibitem{jussim-eccles}
Jussim L, Eccles J.
\newblock Social perception, social stereotypes, and teacher expectations:
  Accuracy and the quest for the powerful self-fulfilling prophecy.
\newblock Advances in Experimental Social Psychology. 1996;28:281--388.

\bibitem{mazur-peer-instruction}
Mazur E.
\newblock Peer Instruction: A User's Manual.
\newblock Prentice Hall; 1996.

\bibitem{porter-multi}
Porter L, Bailey~Lee C, Simon B, Cutts Q, Zingaro D.
\newblock Experience Report: A Multi-classroom Report on the Value of Peer
  Instruction.
\newblock In: Proceedings of the 16th Annual Joint Conference on Innovation and
  Technology in Computer Science Education; 2011. p. 138--142.

\bibitem{porter-success}
Porter L, Guzdial M, McDowell C, Simon B.
\newblock Success in introductory programming: What works?
\newblock Communications of the ACM. 2013;56(8).

\bibitem{nrc-reaching}
Council NR.
\newblock Reaching Students: What Research Says About Effective Instruction in
  Undergraduate Science and Engineering.
\newblock Washington, DC: The National Academies Press; 2015.
\newblock Available from:
  \url{https://www.nap.edu/catalog/18687/reaching-students-what-research-says-about-effective-instruction-in-undergraduate}.

\bibitem{rubin-live-coding}
Rubin MJ.
\newblock The Effectiveness of Live-coding to Teach Introductory Programming.
\newblock In: Proceeding of the 44th ACM Technical Symposium on Computer
  Science Education. SIGCSE '13. New York, NY, USA: ACM; 2013. p. 651--656.
\newblock Available from: \url{http://doi.acm.org/10.1145/2445196.2445388}.

\bibitem{barker-live-coding}
Barker LJ, Garvin-Doxas K, Roberts E.
\newblock What Can Computer Science Learn from a Fine Arts Approach to
  Teaching?
\newblock In: Proceedings of the 36th SIGCSE Technical Symposium on Computer
  Science Education. SIGCSE '05. New York, NY, USA: ACM; 2005. p. 421--425.
\newblock Available from: \url{http://doi.acm.org/10.1145/1047344.1047482}.

\bibitem{willingham-book}
Willingham DT.
\newblock Why don't students like school?: A cognitive scientist answers
  questions about how the mind works and what it means for the classroom.
\newblock John Wiley \& Sons; 2009.

\bibitem{crouch-demo}
Crouch C, Fagen AP, Callan JP, Mazur E.
\newblock Classroom demonstrations: Learning tools or entertainment?
\newblock American Journal of Physics. 2004;72(6):835--838.
\newblock doi:{10.1119/1.1707018}.

\bibitem{miller-demo}
Miller K, Lasry N, Chu K, Mazur E.
\newblock Role of physics lecture demonstrations in conceptual learning.
\newblock Phys Rev ST Phys Educ Res. 2013;9:020113.
\newblock doi:{10.1103/PhysRevSTPER.9.020113}.

\bibitem{hannay}
Hannay JE, Dybå T, Arisholm E, , Sjøberg DIK.
\newblock The effectiveness of pair programming: a meta-analysis.
\newblock Information and Software Technology. 2009;51(7).

\bibitem{mcdowell}
McDowell C, Werner L, Bullock HE, Fernald J.
\newblock Pair Programming Improves Student Retention, Confidence, and Program
  Quality.
\newblock Communications of the ACM. 2006;49(8):90--95.

\bibitem{morrison-subgoals}
Morrison BB, Margulieux LE, Guzdial M.
\newblock Subgoals, Context, and Worked Examples in Learning Computing Problem
  Solving.
\newblock In: Proceedings of the Eleventh Annual International Conference on
  International Computing Education Research. ICER '15. New York, NY, USA: ACM;
  2015. p. 21--29.
\newblock Available from: \url{http://doi.acm.org/10.1145/2787622.2787733}.

\bibitem{morrison-parsons}
Morrison BB, Margulieux LE, Ericson B, Guzdial M.
\newblock Subgoals Help Students Solve Parsons Problems.
\newblock In: Proceedings of the 47th ACM Technical Symposium on Computing
  Science Education. SIGCSE '16. New York, NY, USA: ACM; 2016. p. 42--47.
\newblock Available from: \url{http://doi.acm.org/10.1145/2839509.2844617}.

\bibitem{margulieux}
Margulieux LE, Guzdial M, Catrambone R.
\newblock Subgoal-labeled Instructional Material Improves Performance and
  Transfer in Learning to Develop Mobile Applications.
\newblock In: Proceedings of the Ninth Annual International Conference on
  International Computing Education Research; 2012. p. 71--78.

\bibitem{gick-holyoak}
Gick ML, Holyoak KJ.
\newblock The cognitive basis of knowledge transfer.
\newblock In: Cormier SM, Hagman JD, editors. Transfer of learning:
  Contemporary research and applications. Academic Press; 1987.

\bibitem{guzdial-media}
Guzdial M.
\newblock Exploring hypotheses about media computation.
\newblock In: Proceedings of the Ninth Annual International ACM Conference on
  International Computing Education Research. ICER '13; 2013. p. 19--26.

\bibitem{bouvier-context}
Bouvier D, Lovellette E, Matta J, Alshaigy B, Becker BA, Craig M, et~al.
\newblock Novice Programmers and the Problem Description Effect.
\newblock In: Proceedings of the 2016 ITiCSE Working Group Reports. ITiCSE '16.
  New York, NY, USA: ACM; 2016. p. 103--118.
\newblock Available from: \url{http://doi.acm.org/10.1145/3024906.3024912}.

\bibitem{repenning}
Repenning A.
\newblock Moving beyond syntax: lessons from 20 years of blocks programing in
  {AgentSheets}.
\newblock Journal of Visual Languages and Sentient Systems. 2017;3.
\newblock doi:{10.18293/VLSS2017-010}.

\bibitem{parnin-expertise}
Parnin C, Siegmund J, Peitek N.
\newblock On the Nature of Programmer Expertise.
\newblock In: Psychology of Programming Interest Group Workshop 2017; 2017.

\bibitem{parsons}
Parsons D, Haden P.
\newblock Parsons Programming Puzzles: A Fun and Effective Learning Tool for
  First Programming Courses.
\newblock In: Proceedings of the 8th Australasian Conference on Computing
  Education - Volume 52. ACE '06. Darlinghurst, Australia, Australia:
  Australian Computer Society, Inc.; 2006. p. 157--163.
\newblock Available from:
  \url{http://dl.acm.org/citation.cfm?id=1151869.1151890}.

\bibitem{ericson-ebook}
Ericson B, Moore S, Morrison B, Guzdial M.
\newblock Usability and Usage of Interactive Features in an Online Ebook for CS
  Teachers.
\newblock In: Proceedings of the Workshop in Primary and Secondary Computing
  Education. WiPSCE '15. New York, NY, USA: ACM; 2015. p. 111--120.
\newblock Available from: \url{http://doi.acm.org/10.1145/2818314.2818335}.

\end{thebibliography}
